---
title: "POLI 784 - Problem Set #3"
author: "Nico Cardenas-Miller, Kieran Fitzmason, Otilda Harris"
date: today
format: pdf
editor: visual
---

# Question 1

```{r}
#| echo: false
#| include: false

# Install required packages if you haven't already
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("dagitty")) install.packages("dagitty")
if (!require("ggdag")) install.packages("ggdag")
if (!require("MatchIt")) install.packages("MatchIt")
if (!require("cobalt")) install.packages("cobalt")
if (!require("patchwork")) install.packages("patchwork")
if (!require("marginaleffects")) install.packages("marginaleffects")


# Load packages into environment
library(tidyverse)
library(dagitty)
library(ggdag)
library(MatchIt)
library(cobalt)
library(patchwork)
library(marginaleffects)
```

## Part A

Here, we simulate a dataset that includes variables representing our treatment $D$, outcome of interest $Y$, observed covariates $\mathbf{X}$, and an unobserved covariate $U_1$. The causal relationships between these variables is represented below as a directed acyclic graph (DAG).

### DAG representation of causal relationships

```{r}

# Random seed has an effect on the placement of variables in the DAG plot. 
# Pick one that makes the DAG look pretty. 
set.seed(1200)

# Represent relationship between variables as a DAG
dag1A <- dagify(
  X1 ~ U1,
  D ~ X1 + X2, 
  X3 ~ D,
  Y ~ D + X3 + X2 + U1,
  X4 ~ D + Y, 
  exposure = "D",
  outcome = "Y",
  latent = c("U1")
)

# Styled plot
ggdag_status(dag1A) +
  theme_dag() +
  labs(title = "DAG 1A")
```

### Simulated data

```{r}

# Set random seed for reproducibility
set.seed(42)

## Simulate data

# Specify sample size
N <- 1000

# Specify direct effect of D on Y
tau_DY <- 1.0

# Specify indirect effect of D on Y via X3
tau_DX3 <- 0.5
tau_X3Y <- 1.0

# Calculate direct and total effect of D on Y
tau_direct <- tau_DY
tau_indirect <- tau_DX3*tau_X3Y
tau_total <- tau_direct + tau_indirect

# Initialize dataframe
df <- tibble(U1 = rep(NA,N),
             X1 = NA,
             X2 = NA,
             X3 = NA,
             X4 = NA,
             D = NA,
             Y = NA)

# Populate dataframe
df$U1 <- rbinom(N,1,0.5)
df$X1 <- 1 + 0.5*df$U1 + rnorm(N,0,1)
df$X2 <- rbinom(N,1,0.5)
df$D <- 1.25 + 0.5*df$X1 + 0.25*df$X2 + rnorm(N,0,0.1)
df$X3 <- 1 + tau_DX3*df$D + rnorm(N,0,0.1)
df$Y <- 5 + tau_DY*df$D + tau_X3Y*df$X3 + 0.1*df$X2 + 0.1*df$U1 + rnorm(N,0,0.25)
df$X4 <- -0.5 + 2*df$D + 0.75*df$Y + rnorm(N,0,1)

# Display to console 
head(df)
```

## Part B

Here, we estimate the direct effect of $D$ on $Y$ using linear regression. In order to block the back-door paths between $D$ and $Y$, we must condition on $X_1$ and $X_2$ by including them as variables in our regression model. Because we are interested in isolating the direct effect of $D$ on $Y$ we will also condition on $X_3$, which acts as a mediator. This is analogous to asking: how would changing the value of $D$ by one unit affect $Y$ if we were to hold $X_3$ constant?

We can also use the `dagitty::adjustmentSets` function to verify that this is the correct set of variables to condition on given the structure of our causal graph.

```{r}

# Get the set of variables to condition on when estimating the direct effect of D on Y. 
C_direct <- adjustmentSets(dag1A,effect="direct")
print(paste("Conditioning set:",lapply(C_direct, paste, collapse = ",")))

# Estimate direct effect of D on Y 
model <- lm(Y ~ D + X1 + X2 + X3, data = df)
tau_direct_hat <- as.numeric(model$coefficients["D"])

# Print results to console
print(paste("Estimated direct effect of D on Y:",tau_direct_hat))
print(paste("Actual direct effect of D on Y:",tau_direct))
```

As shown above, the linear model is able to recover an unbiased estimate of the direct effect of $D$ on $Y$ when we condition on $X_1$, $X_2$, and $X_3$. Our estimated coefficient is not the exact same as the true effect due to the finite sample size, but repeated simulation with different random seeds would show that our estimation approach is unbiased in expectation.

## Part C

Here, we estimate the total effect of $D$ on $Y$. We will condition on $X_1$ and $X_2$ to block the back-door paths between $D$ and $Y$ as we did in part B; however, we will not include $X_3$ in our conditioning set because we are interested in capturing how $D$ impacts $Y$ both directly and through its interactions with mediator variables.

```{r}

# Get the set of variables to condition on when estimating the total effect of D on Y. 
C_total <- adjustmentSets(dag1A,effect="total")
print(paste("Conditioning set:",lapply(C_total, paste, collapse = ",")))

# Estimate total effect of D on Y 
model <- lm(Y ~ D + X1 + X2, data = df)
tau_total_hat <- as.numeric(model$coefficients["D"])

# Print results to console
print(paste("Estimated total effect of D on Y:",tau_total_hat))
print(paste("Actual total effect of D on Y:",tau_total))
```

As shown above, the linear model is able to recover an unbiased estimate of the total effect of $D$ on $Y$ when we condition on $X_1$ and $X_2$. Our estimated coefficient is not the exact same as the true effect due to the finite sample size, but repeated simulation with different random seeds would show that our estimation approach is unbiased in expectation.

## Part D

Here, we evaluate the impact of adding a $X_4$ to our conditioning set on estimates of the total effect of $D$ on $Y$.

```{r}

# Print conditioning set
print("Conditioning set: X1,X2,X4")

# Estimate total effect of D on Y 
model <- lm(Y ~ D + X1 + X2 + X4, data = df)
tau_total_hat <- as.numeric(model$coefficients["D"])

# Print results to console
print(paste("Estimated total effect of D on Y:",tau_total_hat))
print(paste("Actual total effect of D on Y:",tau_total))
```

By including $X_4$ in our conditioning set, we have opened up a back-door path between $D$ and $Y$ and introduced bias into our causal effect estimate, which now underestimates the true effect. Previously, this back-door path was blocked because $X_4$ is a collider variable. Overcontrolling for a collider variable in this manner is commonly referred to as "collider bias."

## Part E

Here, we modify the DAG from Part A by (1) removing the direct effect of $D$ on $Y$, (2) removing $X_1$ from the graph, and (3) adding a direct effect of $U_1$ on $D$.

```{r}

# Represent relationship between variables as a DAG
dag1E <- dagify(
  D ~ U1 + X2, 
  X3 ~ D,
  Y ~ X3 + X2 + U1,
  X4 ~ D + Y, 
  exposure = "D",
  outcome = "Y",
  latent = c("U1")
)

# Styled plot
ggdag_status(dag1E) +
  theme_dag() +
  labs(title = "DAG 1E")
```

In the above DAG, the back-door criterion would tell us that in order to estimate the causal effect of $D$ on $Y$, we must condition on $X_2$ and $U_1$. Both of these variables are confounders (i.e., they influence both the treatment and the outcome); however, only $X_2$ is observed, which means that the back-door criterion fails.

In order to estimate the causal effect of $D$ on $Y$, we need to take a different approach known as the front-door criterion. This involves two steps: (1) estimating the effect of $D$ on $X_3$, without conditioning on any other variables and (2) estimating the effect of $X_3$ on $Y$ while conditioning on $D$. Both of these effects will be unbiased because all back-door paths between $D$ and $X_3$ are blocked by colliders, and $D$ blocks all back-door paths between $X_3$ and $Y$. As such, multiplying these two effects together should give us an unbiased estimate of the effect of $D$ on $Y$.

## Part F

Here, we modify the DAG from Part E by adding an additional latent variable $U_2$ that directly influences both $X_2$ and $Y$.

```{r}

# Represent relationship between variables as a DAG
dag1F <- dagify(
  D ~ U1 + X2,
  X2 ~ U2,
  X3 ~ D,
  Y ~ X3 + X2 + U1 + U2,
  X4 ~ D + Y, 
  exposure = "D",
  outcome = "Y",
  latent = c("U1","U2")
)

# Styled plot
ggdag_status(dag1F) +
  theme_dag() +
  labs(title = "DAG 1F")
```

Adding $U_2$ doesn't actually change much: the back-door criterion would still tell us to condition on $X_2$ and $U_1$, which blocks the back-door path passing through $U_2$. As in Part E, the back-door criterion fails since $U_1$ is unobserved, but we can still use the front-door criterion to estimate the causal effect of $D$ on $Y$.

If the new DAG represents the true data-generating process, then this implies that the DAG from Part E is technically not a valid causal DAG—at least not according to the definition given in Chapter 6 of [Hernán and Robins (2020)](https://static1.squarespace.com/static/675db8b0dd37046447128f5f/t/691fb7706ce66332f0b44467/1763686256720/hernanrobins_WhatIf_21nov25.pdf#page=85.00). Their definition states that one of the requirements of a causal graph is that "all common causes, even if unmeasured, of any pair of variables on the graph are themselves on the graph." Because the DAG in Part E omits $U_2$, which is a common cause of $X_2$ and $Y$, it does not meet their definition of a causal graph. However, this is mainly a technical point; even though the DAG in Part E omits $U_2$, it still includes enough relevant variables for us to design a strategy for estimating the effect of $D$ on $Y$.

## Part G

Based on the results of Parts A-F, we can draw the following conclusions:

-   In general, we should always control for confounders since this closes back-door paths.

-   In general, we should never control for colliders as this opens up back-door paths.

-   We should only control for mediators when attempting to isolate direct effects from total effects.

These guidelines assume that our causal DAG is correctly specified and captures all variables that are relevant to the data-generating process (including those that are unobserved). In practice, all models are wrong, and there's always a chance that an important variable or causal relationship was omitted or misspecified. While we can reduce these risks by carefully reviewing the literature and eliciting expert opinion when constructing our DAG, it's impossible to eliminate them completely. As such, it is a good idea to construct multiple DAGs that reflect uncertainties in the structure of the data-generating process. In some cases, it may be possible to design estimation strategies that perform well across multiple representations of the data-generating process.

# Question 2

In this question, we work with data from a study by [Blattman and Annan (2010)](https://chrisblattman.com/documents/research/2010.Consequences.RESTAT.pdf) to evaulate the effects of being abducted by a rebel group on educational outcomes among youth in Northern Uganda.

```{r}

# Load dataset
rm(list = ls())
data <- read_csv("data-1.csv")
```

## Part A

This was an observational study because the treatment assignment—in this case, being abducted by a rebel group—was imposed by circumstances outside the control of the researchers. In order to identify causal effects, this study design makes a crucial assumption known as "strong ignorability," which assumes that that treatment assignment is independent of potential outcomes conditional on a set of observed covariates.

$$
\{ Y_i(1), Y_i(0)\}\perp D_i | \textbf{X}_i 
$$

This assumptions allows us to overcome the fundamental problem of causal inference by using control units with similar covariate values as counterfactuals for treated units. When estimating the ATT using a matching estimator, this allows us to impute the missing potential outcome for each treated unit as a weighted average of the observed outcomes of that unit's matched controls.

For this identification assumption to be plausible, the set of covariates $\textbf{X}$ must include all confounders and variables that affect $D$ and $Y$. This means that $\textbf{X}$ should capture all factors that could potentially bias the estimated effect of abduction on education if not accounted for.

## Part B

Here, we examine the covariate balance of the unadjusted data using the following methods:

1.  Conducting balance tests supported by the `matchit()` function.

2.  Creating a love plot of standardized mean differences (SMDs) in covariates between those assigned to treatment and control.

3.  Plotting the eCDFs of covariates for those assigned to treatment and control.

    ```{r}

    # Match on pre-treatment covariates
    set.cobalt.options(binary = 'std')
    SMD <- matchit(
        abd ~ age + fthr_ed + mthr_ed + hh_fthr_frm + hh_size96 + orphan96,
        data = data,
        method = NULL,
        distance = "mahalanobis",
    )
    summary(SMD)
    love.plot(
        x =SMD,
        thresholds = c(m = 0.1),
        abs = TRUE
        )

    # save names of covariates in vector
    covariates <- names(data)[!names(data) %in% c("abd", "educ")]

    # plot balance in unadjusted data using eCDFs
    plots <- list()
    for (var in covariates) {
      plots[[var]] <- bal.plot(
        x = SMD,
        var.name = var,
        which = "both",
        type = "ecdf"
      ) +
      ggtitle(var)
    }
    plots
    ```

The results suggest that there are differences in the covariate distributions between the abducted and non-abducted groups, as indicated by the SMDs and the love plot. These differences are most pronounced for age, with the treated units being older on average. It makes sense that this variable would differ between treated and control units since the paper states that the rebel group preferred to abduct males in a specific age range. ECDF plots also show that there are some differences in the distributions of certain covariates between the two groups—most notably for age. This suggests that ignorability may not hold before matching, as there may be confounding variables that are not balanced between the groups.

## Part C

Here, we conduct nearest neighbor matching for estimating the ATT using the Mahalanobis distance, at a 1:1 ratio, with no replacement.

```{r}

# Match on pre-treatment covariates
SMD_match <- matchit(
    abd ~ age + fthr_ed + mthr_ed + hh_fthr_frm + hh_size96 + orphan96,
    data = data,
    method = "nearest",
    distance = "mahalanobis",
    ratio = 1,
    replace = FALSE
)
summary(SMD_match)
love.plot(
    x = SMD_match,
    thresholds = c(m = 0.1),
    abs = TRUE
    )

# save names of covariates in vector
covariates <- names(data)[!names(data) %in% c("abd", "educ")]

# plot balance in unadjusted data using eCDFs
plots <- list()
for (var in covariates) {
  plots[[var]] <- bal.plot(
    x = SMD_match,
    var.name = var,
    which = "both",
    type = "ecdf"
  ) +
  ggtitle(var)
}
plots
```

Because our original dataset contains 462 treated units but only 279 control units, the above matching scheme results in 183 treated units being discarded. Even among the 279 pairs of treated and control units in our matched dataset (n=558), many of the matches are likely to be of poor quality due to the small number of controls. We see this reflected in the love plots and eCDFs, which show that our matching procedure has actually led to more imbalance in many covariates. In this case, matching does not appear to have made our ignorability assumptions more plausable, and may even have made things worse.

When relying on low-dimensional summaries to assess balance, we can potentially miss important information on distributional differences that a simple difference-in-means or other similar test would miss. Empirical CDFs and related statistical tests (e.g., [KS test](https://en.wikipedia.org/wiki/Kolmogorov%E2%80%93Smirnov_test)) are a better way of comparing distributions, but typically compare the marginal distributions of covariates separately from one another. Joint distribution plots may be another useful tool for assessing covariate balance that can potentially help to identify differences in the joint distribution of covariates.

## Part D

Here, we conduct nearest neighbor matching for estimating the ATT using the CBPS distance metric, a 5:1 ratio, replacement, and pruning of treatment and control units with no common support.

```{r}

# Match on pre-treatment covariates
SMD_match_cbps <- matchit(
    abd ~ age + fthr_ed + mthr_ed + hh_fthr_frm + hh_size96 + orphan96,
    data = data,
    method = "nearest",
    distance = "CBPS",
    ratio = 5,
    replace = TRUE,
# prune treatment and control units with no common support
    discard = "both"
)
summary(SMD_match_cbps)
love.plot(
    x = SMD_match_cbps,
    thresholds = c(m = 0.1),
    abs = TRUE
    ) 

# save names of covariates in vector
covariates <- names(data)[!names(data) %in% c("abd", "educ")]

# plot balance in unadjusted data using eCDFs
plots <- list()
for (var in covariates) {
  plots[[var]] <- bal.plot(
    x = SMD_match_cbps,
    var.name = var,
    which = "both",
    type = "ecdf"
  ) +
  ggtitle(var)
}
plots
```

The above matching procedure resulted in only 2 treated units and 4 control units being discarded, meaning that the size of our matched dataset (which has an effective sample size of 735) is similar to the size of our original dataset (n=741). The love plots and eCDFs also indicate that the matching procedure has substantially improved the amount of covariate balance. Overall, these results suggest that our ignorability assumption should be plausible for the matched dataset, assuming that we have measured all the relevant covariates. We can now estimate the ATT using the code below.

```{r}

# create matched data
matched_data <- match.data(SMD_match_cbps)

# estimate ATT using marginaleffects
model_cbps <- lm(
  educ ~ abd + age + fthr_ed + mthr_ed + hh_fthr_frm + hh_size96 + orphan96,
  data = matched_data,
  weights = weights
)

att_estimate <- marginaleffects::avg_comparisons(model_cbps, variables = "abd")
att_estimate
```

The above estimate of the ATT implies that youth who were abducted by the rebel group lost 0.47 years of education on average as a result of their abduction. This effect size is smaller than the one reported in the paper, but this may not be an apples-to-apples comparison since the paper estimates the ATE instead of the ATT and adjusts for other covariates not included in our model.
