---
title: "POLI 784 - Problem Set #6"
author: "Emily Arnsberg, Kieran Fitzmason, Otilda Harris"
date: today
format: pdf
editor: visual
---

```{r}
# Load packages
library("tidyverse")
library("plm")
library("sandwich")
library("panelView")
library("fixest")
library("fect")
library("sensemakr")
library("glue")
```

# Question 1

In this question, we analyze data from a study by [Sances (2016)](https://joeornstein.github.io/pols-4641/readings/Sances%20-%202016%20-%20The%20Distributional%20Impact%20of%20Greater%20Responsiveness%20Evidence%20from%20New%20York%20Towns.pdf), which evaluated the effect of a town switching from an elected to an appointed tax assessor on the frequency of property tax reassessments in New York state during the 1980s through early 2000s.

## Part A

```{r}
# Load data
data <- read.csv("data1-2.csv")

# Print number of unique units
num_units <- length(unique(data$swis_code))
glue("Number of units: {num_units}")

# Print number of unique periods
num_periods <- length(unique(data$year))
glue("Number of periods: {num_periods}")

# Visualize treatment status using panelview
panelview(data=data,
          formula=reassessed~appointed,
          index=c("swis_code","year"))
```

## Part B

Here, we estimate the effect of switching to an appointed assessor using a classic 2x2 DiD design that compares towns that switched in 1999 to those that never switched.

```{r}

# Filter data to include only towns that switched in 1999 (treated)
# and those that never switched (controls)
dat_DID <- data[data$cohort %in% c(1,14),]

# Create a treated/control indicator
dat_DID$treated <- as.numeric(dat_DID$cohort == 14)

# Create a pre/post treatment indicator
dat_DID$post_treatment <- as.numeric(dat_DID$year >= 1999)

# Fit classic 2x2 DiD regression
mod_1B <- lm(reassessed ~ treated * post_treatment, data = dat_DID)
summary(mod_1B)

# Extract ATT estimate
ATT_1B <- mod_1B$coefficients["treated:post_treatment"]

# Print to console
glue("Classic 2x2 DiD ATT estimate: {round(ATT_1B,6)}")
```
